Immutable Source-Mapped Abstract Syntax Tree: \\ A Design Pattern for Refactoring Engine APIs

### 2013-10-01 paper.tex
Final revisions

% TODO: Version with [draft] is for draft mode only
% Also comment/uncomment \today in subtitle when in draft mode
\documentclass[prodmode]{acmlarge}
%\documentclass[draft]{acmlarge}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{url}
\usepackage{datetime} % for \today
\usepackage{listings} % for lstlisting environment
\usepackage{placeins} % for \FloatBarrier
\newcommand{\ttt}[1]{\texttt{#1}}
\newcommand{\code}[1]{\texttt{#1}}

@@

\begin{abstract}
Many automated refactoring tools provide an extensibility mechanism, which
allows third parties to contribute new refactorings.  However, implementing
nontrivial refactorings requires deep syntactic knowledge about the program
being refactored.  Refactoring tools almost always use abstract syntax trees
(ASTs) to describe the structure of source code.  Therefore, it may be
desirable to expose these trees via an API.  This pattern describes the common
attributes that ASTs typically exhibit in order to be useful for implementing
nontrivial refactorings while, at the same time, maintaining the
characteristics of good API design.  This pattern is used in the Eclipse Java
Development Tools' refactoring API as well as the Microsoft ``Roslyn'' CTP,
among other uses.
\end{abstract}

@@

If an IDE provides an application programming interface (API) that allows third
parties to contribute new refactorings, then it is often helpful for that API
to provide access to ASTs as well.  This pattern considers ASTs exposed via
refactoring engine extensibility APIs, identifying the common elements
necessary to provide both a clean API and a data structure suitable for
implementing nontrivial refactorings.

@@

\item Clients must be able to use the API to manipulate source code in a way
that is minimally invasive: refactorings that only affect a small, isolated
region of the source code should not change any source code outside that
region.  This includes preserving the program's comments and formatting.

@@

\begin{wrapfigure}{r}{3in} % r is right, R is float right
% Original image is 4.624 x 10.651 in
\includegraphics[scale=0.62]{jdt-ast.eps}
%\caption{Visualization of the Eclipse Java Development Tools' abstract syntax
%tree for the following program: \\
%\ \\
%\centerline{\parbox{2in}{\ttt{\noindent public class Example \{ \\
%\hspace*{.2em} public int FIVE = 5; \\
%\}
%}}} \\
%\ \\
%The highlighted node---a \textit{VariableDeclarationFragment}---corresponds to
%the eight-character substring \ttt{FIVE = 5}, which begins with the 36th
%character in the source code.
%}
\caption{Visualization of an AST from Eclipse JDT.}
\label{fig:jdt-ast}
\end{wrapfigure}

@@

\vspace*{1em}
\noindent (This visualization is produced by the Eclipse Java Development Tools
(JDT), which are well known for their refactoring capabilities.)

Notice how deeper levels of the tree describe progressively more fine-grained
syntactic structures.  The root node represents the entire Java source file.
The highlighted node---a \textit{VariableDeclarationFragment}---represents
\ttt{FIVE = 5}, while its child nodes represent its constituent parts: the
identifier \ttt{FIVE} and the numeric literal \ttt{5}.  These are leaf nodes,
since these tokens (like keywords, identifiers, and numeric literals) represent
the smallest syntactic units in the language.

@@

Unfortunately, while a mutable AST can provide an elegant means of manipulating
source code, that are other issues that can make mutable ASTs problematic from
an API design perspective.  Some notable issues include the following.
\begin{itemize}
\item \textit{AST nodes contain source location information.}  After the
structure of the AST is changed, what does the source location information in
each node represent: the ``old'' source location (in the original code) or the
``new'' location (after it has been modified)?  If it represents the old
location, what should the source location be when new nodes are inserted into
the tree?  If it represents the new location, then every time a change is made
to the tree, source location information must be updated for every node in the
tree, which may be expensive \dots or else a clever implementation (e.g.,
computation from relative offsets) must be used instead.
\item \textit{AST nodes may be removed from the tree, leaving them orphaned.}
Suppose a client traverses the AST, storing pointers to some ``interesting''
nodes in a collection.  Then, an ancestor of one of those nodes is removed from
the AST.  This will leave the client in a difficult-to-debug situation where
his collection contains pointers to nodes that no longer exist.
\item \textit{Semantic information may be invalidated.}  Refactorings often
need to determine, for a particular use of a variable, what declaration that
use corresponds to.  For example, in Java, if a field and a local variable were
both named \ttt{x}, then in the expression \ttt{this.x = x}, the first \ttt{x}
would refer to the field, while the second would refer to the local variable.
However, if the local variable declaration were removed, the second \ttt{x}
would refer to the field instead.  This can be confusing to clients, since
information collected about the program at one point in time may no longer be
true after the tree's structure changes.
\item \textit{It may be possible to mutate the AST so that it does not
represent a legal program.}  It might even be possible to mutate the AST so
that it is no longer a tree at all.  In Java, classes can contain other
classes.  What happens if a client tries to make a class an inner class of
itself?  (This could turn the tree into a cyclic graph.)  The API for modifying
the AST's structure could try to prevent such things, but attempting to detect
and prevent every possible mistake would be difficult and potentially
expensive.
\item \textit{Mutability can be problematic in multithreaded code.}
\end{itemize}

Mutable ASTs were used in the original Smalltalk Refactoring
Browser~\cite{roberts97refactoring}, as well as in
CRefactory~\cite{garrido05program} and
Photran (details of Photran's implementation are discussed
in~\cite{sle2008}).  Many compilers use mutable ASTs internally.  However, these
tools have one thing in common: the mutable AST is not accessible via any
public API.  It is intended for internal use only, by developers who should be
better equipped to handle all of the caveats.

Guidelines for API design are discussed in detail in Jaroslav Tulach's
\textit{Practical API Design}~\cite{tulach08practical}.  A presentation by
Joshua Bloch~\cite{bloch} offers excellent advice as well.  Both authors
suggest minimizing the amount of mutability in an API.  After all, mutable
objects tend to have simpler interfaces, and they are thread-safe.  Both
authors also emphasize that APIs should be easy to learn (even for programmers
who do not read all of the API documentation) and difficult for clients to
misuse.  Given these guidelines, and the list of liabilities for mutable ASTs
given above, it seems preferable from the perspective of someone attempting to
publish an API to avoid mutable ASTs.  But this raises a different question: If
ASTs are immutable, what API should be provided for clients to manipulate
source code?

@@

The situation gets a bit worse when changes are made to nested constructs.
Recall the example from Section~\ref{sec:Problem}, which added braces to two
\textit{if} statements, one nested inside the other.  Suppose the inner
\textit{if} statement were modified first: Adding braces to it would change the
position at which the closing brace for the outer \textit{if} statement would
need to be added.  Likewise, if braces were added to the outer \textit{if}
statement first, then this would shift the location of the inner \textit{if}
statement.  In other words, when refactorings affect nested language
constructs, offset/length computations can become complex.

@@

One option is to provide an API that allows clients to specify what changes
they want to make in terms of AST nodes.  This API can then translate these
into string edits.  For example, an API might offer a \textit{Rewriter} class.
The client would inform the \textit{Rewriter} that it wants to effectively
delete a particular \textit{FieldDeclaration} node from the AST:
\textit{rewriter.delete(field)}.  When it is ready to manipulate the source
code, it delegates this task to the rewriter:
\textit{rewriter.rewrite(sourceCode)}.

@@

Immutable, source-mapped ASTs are used by the Eclipse Java Development Tools
(JDT) and C/C++ Development Tools (CDT), the Microsoft ``Roslyn'' Community
Technology Preview (CTP), Clang, and the original refactoring engine in Apple
Xcode.  However, while all use immutable ASTs with source mapping information,
the actual rewriting APIs vary substantially.

@@

@@

>>>>

% TODO: Version with [draft] is for draft mode only
% Also comment/uncomment \today in subtitle when in draft mode
\documentclass[prodmode]{acmlarge}
%\documentclass[draft]{acmlarge}
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{url}
\usepackage{datetime} % for \today
\usepackage{listings} % for lstlisting environment
\usepackage{placeins} % for \FloatBarrier
\newcommand{\ttt}[1]{\texttt{#1}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\jb}[1]{\fbox{\texttt{#1}\rule[-1pt]{0pt}{.9em}}}
\newcommand{\kb}[1]{\mbox{\texttt{\small #1}\rule{7.3pt}{0pt}\rule[-1pt]{0pt}{.9em}}}
\newcommand{\lb}[1]{\mbox{\texttt{\small #1}\rule{3.2pt}{0pt}\rule[-1pt]{0pt}{.9em}}}

@@

\begin{abstract}
Many automated refactoring tools provide an extensibility mechanisms that
allows third parties to contribute new refactorings.  However, implementing
nontrivial refactorings requires deep syntactic knowledge about the program
being refactored.  Refactoring tools almost always use abstract syntax trees
(ASTs) to describe the structure of source code.  Therefore, it may be
desirable to expose these trees via an API.  This pattern describes the common
attributes that ASTs typically exhibit in order to be useful for implementing
nontrivial refactorings while, at the same time, maintaining the
characteristics of good API design.  This pattern is used in the Eclipse Java
Development Tools' refactoring API as well as the Microsoft ``Roslyn'' CTP,
among other uses.
\end{abstract}

@@

If an IDE provides an application programming interface (API) that allows third
parties to contribute new refactorings, then it is often helpful for that API
to provide access to ASTs as well.  This pattern considers ASTs exposed by such
APIs, identifying the common elements necessary to provide both a clean API and
a data structure suitable for implementing nontrivial refactorings.

@@

\item Clients must be able to use the API to manipulate source code in a way
that is minimally invasive: refactorings that only affect a small, isolated
region of the source code should not change any source code outside that
region.  This includes preserving the program's comments and
formatting~\cite{sommerlad08retaining}.

@@

\begin{wrapfigure}{r}{3in} % r is right, R is float right
% Original image is 4.624 x 10.651 in
\includegraphics[scale=0.62]{jdt-ast.eps}
%\caption{Visualization of the Eclipse Java Development Tools' abstract syntax
%tree for the following program: \\
%\ \\
%\centerline{\parbox{2in}{\ttt{\noindent public class Example \{ \\
%\hspace*{.2em} public int FIVE = 5; \\
%\}
%}}} \\
%\ \\
%The highlighted node---a \textit{VariableDeclarationFragment}---corresponds to
%the eight-character substring \ttt{FIVE = 5}, which begins with the 36th
%character in the source code.
%}
\caption{Visualization of an AST produced by the Eclipse Java Development Tools
(JDT).}
\label{fig:jdt-ast}
\end{wrapfigure}

@@

\vspace*{1em}
Notice how deeper levels of the tree describe progressively more fine-grained
syntactic structures.  The root node represents the entire Java source file.
The highlighted node---a \textit{VariableDeclarationFragment}---represents
\ttt{FIVE = 5}, while its child nodes represent its constituent parts: the
identifier \ttt{FIVE} and the numeric literal \ttt{5}.  These are leaf nodes,
since these tokens (like keywords, identifiers, and numeric literals) represent
the smallest syntactic units in the language.

@@

Unfortunately, while a mutable AST can provide an elegant means of manipulating
source code, that are other issues that can make mutable ASTs problematic from
an API design perspective.  Some notable issues include the following.
\begin{itemize}
\item \textit{AST nodes contain source location information.}  After the
structure of the AST is changed, what does the source location information in
each node represent: the ``old'' source location (in the original code) or the
``new'' location (after it has been modified)?  If it represents the old
location, what should the source location be when new nodes are inserted into
the tree?  If it represents the new location, then every time a change is made
to the tree, source location information must be updated for every node in the
tree, which may be expensive \dots or else a clever implementation (e.g.,
computation from relative offsets) must be used instead.
\item \textit{AST nodes may be removed from the tree, leaving them orphaned.}
Suppose a client traverses the AST, storing pointers to some ``interesting''
nodes in a collection.  Then, an ancestor of one of those nodes is removed from
the AST.  This will leave the client in a difficult-to-debug situation where
his collection contains pointers to nodes that no longer exist in the AST.
\item \textit{Semantic information may be invalidated.}  Refactorings often
need to determine, for a particular use of a variable, what declaration that
use corresponds to.  For example, in Java, if a field and a local variable were
both named \ttt{x}, then in the expression \ttt{this.x = x}, the first \ttt{x}
would refer to the field, while the second would refer to the local variable.
However, if the local variable declaration were removed, the second \ttt{x}
would refer to the field instead.  This can be confusing to clients, since
information collected about the program at one point in time may no longer be
true after the tree's structure changes.
\item \textit{It may be possible to mutate the AST so that it does not
represent a legal program.}  It might even be possible to mutate the AST so
that it is no longer a tree at all.  In Java, classes can contain other
classes.  What happens if a client tries to make a class an inner class of
itself?  (This could turn the tree into a cyclic graph.)  The API for modifying
the AST's structure could try to prevent such things, but attempting to detect
and prevent every possible mistake would be difficult and potentially
expensive.
\item \textit{Mutability can be problematic in multithreaded code.}
\end{itemize}

Mutable ASTs were used in the original Smalltalk Refactoring
Browser~\cite{roberts97refactoring}, as well as in
CRefactory~\cite{garrido05program} and Photran~\cite{sle2008}.  Many compilers
use mutable ASTs internally.  However, these tools have one thing in common:
the mutable AST is not accessible via any public API.  It is intended for
internal use only, by developers who should be better equipped to handle all of
the caveats.

Guidelines for API design are discussed in detail in Jaroslav Tulach's
\textit{Practical API Design}~\cite{tulach08practical}.  A presentation by
Joshua Bloch~\cite{bloch} offers excellent advice as well.  Both authors
suggest minimizing the amount of mutability in an API.  After all, immutable
objects tend to have simpler interfaces, and they are thread-safe.  Both
authors also emphasize that APIs should be easy to learn (even for programmers
who do not read all of the API documentation) and difficult for clients to
misuse.  Given these guidelines, and the list of liabilities for mutable ASTs
given above, it seems preferable from the perspective of someone attempting to
publish an API to avoid mutable ASTs.  But this raises a different question: If
ASTs are immutable, what API should be provided for clients to manipulate
source code?

@@

The situation gets a bit worse when changes are made to nested constructs.
Recall the example from Section~\ref{sec:Problem}, which added braces to two
\textit{if} statements, one nested inside the other.  Suppose the outer
\textit{if} statement were modified first: Adding braces to it would change the
position at which the closing brace for the inner \textit{if} statement would
need to be added.  Consider a somewhat simplified example.  Pictorially:

\begin{center}
\jb{i}\jb{f}\jb{(}\jb{1}\jb{)}\jb{ }%
\jb{i}\jb{f}\jb{(}\jb{2}\jb{)}\jb{ }%
\jb{s}\jb{(}\jb{)}\jb{;}\\
\lb{\,}\kb{0}\kb{1}\kb{2}\kb{3}\kb{4}\kb{5}\kb{6}\kb{7}\kb{8}\lb{9}%
\lb{10}\lb{11}\lb{12}\lb{13}\lb{14}\lb{15} \\
$\Downarrow$ \rule[-0.5em]{0pt}{1.2em} \\
\jb{i}\jb{f}\jb{(}\jb{1}\jb{)}\jb{ }%
\jb{\{}\jb{ }%
\jb{i}\jb{f}\jb{(}\jb{2}\jb{)}\jb{ }%
\jb{s}\jb{(}\jb{)}\jb{;}%
\jb{ }\jb{\}}\\
\lb{\,}\kb{0}\kb{1}\kb{2}\kb{3}\kb{4}\kb{5}\kb{6}\kb{7}\kb{8}\lb{9}%
\lb{10}\lb{11}\lb{12}\lb{13}\lb{14}\lb{15}\lb{16}\lb{17}\lb{18}\lb{19} \\
\end{center}

\noindent
The body of the inner \textit{if} statement begins at offset~12 in the original
code.  But after adding braces to the outer \textit{if} statement, the body of
the inner \textit{if} statement is shifted to offset~14.

%Likewise, if braces were added to the inner \textit{if} statement first, then
%this would shift the location of the closing brace for the outer \textit{if}
%statement.
In other words, when refactorings affect nested language constructs,
offset/length computations can become complex, since the edits for one
construct may interact with the edits for another construct.

@@

One option is to provide an API that allows clients to specify what changes
they want to make in terms of AST nodes.  This API can then translate these
into string edits.  For example, an API might offer a \textit{Rewriter} class.
The client would inform the \textit{Rewriter} that it wants to effectively
delete a particular \textit{FieldDeclaration} node from the AST:
\textit{rewriter.delete(field)}.  When it is ready to manipulate the source
code, it delegates this task to the rewriter:
\textit{rewriter.rewrite(sourceCode)}.  This is illustrated in
Figure~\ref{fig:seq-diagram}.  The Eclipse Java and C/C++ Development
Tools~\cite{jdt-astrewrite,cdt-refactoring} take this approach
(with slightly different class and method names).

\begin{figure}[!bt]
\begin{center}
\includegraphics[width=\textwidth]{seq-diagram.eps}
\end{center}
\caption{Sequence diagram illustrating use of a Rewriter to modify source code.}
\label{fig:seq-diagram}
\end{figure}

@@

Immutable, source-mapped ASTs are used by the Eclipse Java Development Tools
(JDT) and C/C++ Development Tools (CDT), as well as the Scala IDE for Eclipse,
the Microsoft ``Roslyn'' Community Technology Preview (CTP), Clang, and the
original refactoring engine in Apple Xcode.  However, while all use immutable
ASTs with source mapping information, the actual rewriting APIs vary
substantially.

@@

Refactoring support in the Scala IDE for
Eclipse~\cite{scalaide,scalarefactoring} takes a functional approach,
implementing refactorings as tree transformations.  Source generation is
handled separately: one printer uses position information to retain the
formatting of the original code, while another pretty-prints new nodes that
have been added to the AST~\cite{stocker10scala}.

@@

\textit{The author would like to thank Peter Sommerlad for the invaluable
feedback and suggestions he provided as the shepherd for this paper.}
